---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown for Course Project 1  
Assignment submitted by Miranda L. Gardner  

### Read csv file in and inspect data

```{r}
#READ CSV AND EXAMINE STRUCTURE 
all_data <- read.csv("activity.csv", stringsAsFactors = F, header = T)
str(all_data)

```



### What is the mean number of steps taken daily? 
Note: Ignore missing data reported as NA 

```{r}
#LOAD LIBRARY FOR PIPES, GROUPING AND SUMMARIZE FOR TOTAL STEPS, MEAN STEPS AND MEDIAN BY DAY
library(dplyr)
library(knitr)

df1 <- all_data %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps, na.rm = T), 
            mean_steps = mean(steps, na.rm = T), 
            median_steps = median(steps, na.rm = T))
head(df1)
```


#### Histogram of Total Number of Steps Taken Daily

```{r}
#LOAD LIBRARY TO CREATE HISTOGRAM
library(ggplot2)
ggplot(df1, aes(x= total_steps)) + geom_histogram(color = "black", fill = "magenta") +
  labs(x = "Total Number of Steps Taken Per Day", y = "Number of Days Total Steps Taken") +
  theme_bw() +
  theme(text = element_text(family = "sans", face = "bold", color = "black"), 
        axis.title = element_text(size = 14)
        ) 
```



#### Report the Mean and Median Number of Steps Taken Per Day  

The mean number of steps taken per day:
```{r}
df2 <- df1 %>%
  select(date, mean_steps)
as.data.frame(df2)
```


The median number of steps taken per day:
```{r}
df3 <- df1 %>%
  select(date, median_steps)
as.data.frame(df3)
```


### What is the average daily activity pattern?

```{r}
#GROUP DATA BY INTERVAL AND CALC MEAN NUMBER OF STEPS

df4 <- all_data %>%
  group_by(interval) %>%
  summarize(mean_steps = mean(steps, na.rm = T))
df4
#FIND MAX MEAN STEPS TO FIND THE INTERVAL THAT MAX OCCURS
max <- max(df4$mean_steps)
max_interval <- df4$interval[df4$mean_steps == max]
max_rd = round(max, digits = 3)
#PLOT DATA
 ggplot(df4, aes(x=interval, y=mean_steps)) +
  geom_line(col = "red") + 
   labs(x = "Daily Time Interval", y = "Average number of steps taken", 
        title  = "Average Number of Steps Taken Over Interval") +
   theme_classic() +
   theme(text = element_text(family = "sans", color = "black", face = "bold"), 
         plot.title = element_text(hjust = 0.5)
         )
```

#### Time Interval for Maximum Activity
The time interval that, on average, the maximum number of  `r max_rd ` steps occurs is: `r max_interval`


### Imputing Missing Values

```{r}
#ADD UP TOTAL NUMBER OF MISSING VALUES
num_na <- sum(is.na(all_data))
```

#### Number of Missing Values
The total number of missing values is `r num_na`

```{r}
#FIND INTERVAL MATCH FROM df4 TO all_data & USE mean_steps TO REPLACE NA
impute_match <- df4$mean_steps[match(all_data$interval, df4$interval)]
```


```{r}
#CREATE A NEW DATASET BY REPLACING NAs WITH MEAN FROM THE INTERVAL 
all_data_imputed <- transform(all_data, steps = ifelse(is.na(all_data$steps), yes = impute_match, no = all_data$steps))
```


```{r}
#GROUP DATA BY DATE, SUM DATA & CREATE HISTOGRAM OF IMPUTED DATA
imp_data <- all_data_imputed %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps, na.rm = T), 
            mean_steps = mean(steps, na.rm = T), 
            median_steps = median(steps, na.rm = T))
head(imp_data)

ggplot(imp_data, aes(x= total_steps)) + 
  geom_histogram(color = "black", fill = "purple4") +
  labs(x = "Total Number of Steps Taken Per Day", y = "Number of Days Total Steps Taken") +
  theme_bw() +
  theme(text = element_text(family = "sans", face = "bold", color = "black"), 
        axis.title = element_text(size = 14)
        ) 

```

#### Report the Mean and Median Number of Steps Taken Per Day Post-Imputation

The mean number of steps taken per day:
```{r}
imp_data2 <- imp_data %>%
  select(date, mean_steps)
as.data.frame(imp_data2)
```


The median number of steps taken per day:
```{r}
imp_data3 <- imp_data %>%
  select(date, median_steps)
as.data.frame(imp_data3)
```


The mean and median from the imputed dataset differs from the original calculations but only on the dates where 'NA' was reported for the full day. There are still quite a bit of 0's (no activity) that are further influencing the data.

### Are there differences in activity patterns between weekdays and weekends?

```{r}
#ADD NEW COLUMN WEEKDAY TO TAG DATE AS WD OR WEN
all_data_imputed$day_of_week <- weekdays(as.Date(all_data_imputed$date))
all_data_imputed$day_type <- ifelse (all_data_imputed$day_of_week == "Saturday" | 
                                       all_data_imputed$day_of_week =="Sunday",
                                     "weekend", "weekday")
#GROUP DATA BY INTERVAL
activity_by_date <- aggregate(steps~interval + day_type, all_data_imputed, mean)

#LINE SERIES PLOT - FACET WRAP ON day_type

ggplot(activity_by_date, aes(x = interval, y = steps, color = day_type)) +
  geom_line() +
  facet_wrap(~day_type, ncol = 1, nrow = 2) +
  scale_color_manual(values = c("red", "gray60")) +
   labs(x = "Daily Time Interval", y = "Average number of steps taken",
        title  = "Average Number of Steps By Type of Day", color = "Type of Day") +
   theme_bw() +
   theme(text = element_text(family = "sans", color = "black", face = "bold"),
         plot.title = element_text(hjust = 0.5)
         )
```